Denotational Semantics of General Payment Primitives, and Its Payment System

### 2022-09-28 packages/spec-haskell/yellowpaper/Paper.tex
subsection Functional Reactive Programming drafted

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document Preamble
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[a4paper,10pt]{report}

%% \usepackage[margin=0.5in,nomarginpar]{geometry}

%% Base Folder for Inputs
\makeatletter
\def\input@path{{../tex/}}
\makeatother

%% Import Packages
\usepackage{hyperref}
\usepackage{fvextra}
\usepackage{csquotes}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{graphicx}

%% Formattings

\renewcommand{\mkbegdispquote}[2]{\itshape}

%% Literate Haskell Macros

%%% include common formatting
\input{lhsfmt.tex}
%%% for ignoring some code
\long\def\ignore#1{}
%%% for code
\newminted[code]{haskell}{}
\newminted[spec]{haskell}{}

%% Document Meta Data

@@

\usepackage[
    backend=biber,
    style=alphabetic,
]{biblatex}
\addbibresource{../tex/Biblio.bib}

@@

\begin{itemize}
\item U is the set of monetary units.
\item $\nu : U \rightarrow \mathbb{N}$ is the value function defining the value $\nu(u)$ of every value unit u. The set
    $\mathbb{N}$ is the set of all natural numbers, but instead, we can use any set of numerals that is totally ordered
    (e.g.\ integers, real numbers).
\item $\beta : U \rightarrow \mathfrak{B}$ is the bearer function defining the bearer $\beta(u)$ of a unit. The set
    $\mathfrak{B}$ is the set of possible bearers. The bearer is usually a legal construction defining any type of legal
    entity, such as a person, a family, a company, a state institution, etc.
\end{itemize}

@@

\begin{minted}[escapeinside=||,mathescape=true]{haskell}
    somePaymentPrimitive :: ( Timestamp t
                            , MoneyDistribution md
                            , SharedContext ctx
                            , PaymentInstruction args
                            , Monoid md, Monoid ctx
                            )
                         => md |$\otimes$| ctx -> args -> t -> md |$\otimes$| ctx
\end{minted}

@@

\begin{itemize}
\item \textit{Token and its Accounts} - E.g. bank currency accounts. Each token is its own centralized execution
    environment, bearers access their monetary value through their accounts, and execute financial contracts through the
    token.
\item \textit{Note} - E.g. federal reserve notes, bills, coins and Bitcoin UTXOs, etc. The execution environment is
    independent of the notes, but it needs notes to complete the execution of financial contracts.
\end{itemize}

@@

After FRP got more adoption, it evolved to some variations that support discrete semantics, and some variations better
suited more for interactive systems. In this paper we will stick to and revisit the original formulation use in
\cite{elliott1997functional} where we will draw inspirations from.

@@

We also want a formal and precise specification of payment primitives. The title of the paper mentions denotational
semantics, it is a \textit{compositional} style for precisely specifying the meanings of languages, invented by
Christopher Strachey and Dana Scott in the 1960s (\cite{scott1971toward}). Conal Elliott proposed that denotational
semantics can also be applied to \textit{data types} in programming \footfullcite[][section 2, denotational semantics
    and data types]{Elliott2009-type-class-morphisms-TR}. We will revisit this formulation too.

@@

\section{Complexity Argument}

@@

\section{Constraint in Consensus}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document Setup
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% This document is processed in an isolated clean folder at the same level.
\makeatletter
\def\input@path{{../tex/}}
\makeatother
%%% Include the preamble at choice during processing.
\input{preamble.tex}
%%% Bibliography system setup using biblatex.
\usepackage[
    backend=biber,
    style=alphabetic,
]{biblatex}
\addbibresource{../tex/Biblio.bib}
%%% Macros used.
\long\def\ignore#1{} % for ignoring some code
\newcommand{\eg}{\textit{e}.\textit{g}.}
\newcommand{\ie}{\textit{i}.\textit{e}.}
\newcommand{\etal}{\textit{et al}.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document Meta Data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\begin{itemize}
\item U is the set of monetary units.
\item $\nu : U \rightarrow \mathbb{N}$ is the value function defining the value $\nu(u)$ of every value unit u. The set
    $\mathbb{N}$ is the set of all natural numbers, but instead, we can use any set of numerals that is totally ordered
    (\eg\ integers, real numbers).
\item $\beta : U \rightarrow \mathfrak{B}$ is the bearer function defining the bearer $\beta(u)$ of a unit. The set
    $\mathfrak{B}$ is the set of possible bearers. The bearer is usually a legal construction defining any type of legal
    entity, such as a person, a family, a company, a state institution, etc.
\end{itemize}

@@

\begin{minted}[escapeinside=||,mathescape=true]{haskell}
    somePaymentPrimitive :: ( Timestamp t
                            , MoneyDistribution md
                            , SharedContext ctx
                            , PaymentInstruction args
                            , Monoid md, Monoid ctx
                            )
                         => md |$\times$| ctx -> args -> t -> md |$\times$| ctx
\end{minted}

@@

\begin{itemize}
\item \textit{Token and its Accounts} - \eg\ bank currency accounts. Each token is its own centralized execution
    environment, bearers access their monetary value through their accounts, and execute financial contracts through the
    token.
\item \textit{Note} - \eg\ federal reserve notes, bills, coins and Bitcoin UTXOs, etc. The execution environment is
    independent of the notes, but it needs notes to complete the execution of financial contracts.
\end{itemize}

@@

After FRP got more adoption, it evolved to some variations that support discrete semantics, and some variations better
suited more for interactive systems. In this paper we will stick to and revisit the basic constructions of the original
formulation used in \cite{elliott1997functional} where we will draw inspirations from.

\paragraph{Temporal Modeling and Behaviors}

Values that vary over continuous time are called \textit{behaviors}. They are first-class values, and are built up
compositionally. In modern payment systems, we should want the monetary values of monetary units capable of varying over
continuous time. The semantic function of $\alpha$-behaviors produces the value of type $\alpha$ of a behavior at a given
time:

\begin{equation}
    at : Behavior_{\alpha} \rightarrow Time \rightarrow \alpha
\end{equation}

\paragraph{Event Modeling}

Like behaviors, \textit{events} are first-class values too. The semantic function of $\alpha$-event describes the time and
information associated with an \textit{occurrence} of the event:

\begin{equation}
    occ : Event_{\alpha} \rightarrow Time \times \alpha
\end{equation}

In modern payment systems, the payment primitives executed in payment execution environments are one type of
events. More types of events can be read from the original paper \footfullcite[][section 2.3 Semantics of
    Events]{elliott1997functional}.

\paragraph{Reactivity}

They key to modeling the payment execution environment using FRP is the \textit{reactivity}, which makes behaviors
reactive. Specifically, the behavior \textit{b untilB e} exhibits b's behavior until \textit{e} occurs, and then swiches
to a new behavior encoded in \textit{e}:

\begin{equation}
    \begin{split}
    &untilB : Behavior_{\alpha} \rightarrow Event_{Behavior_{\alpha}} \rightarrow Behavior_{\alpha} \\
    &at\ [\![b\ until\ B]\!]\ t = if\ t\ \leq\ t_{e}\ then\ at\ [\![b]\!]\ t\ else\ at\ [\![b']\!]\ t \\
    &\qquad where (t_e, b') = occ[\![e]\!]
    \end{split}
\end{equation}

Note that $[\![.]\!]$ is the denotational semantics notation to be introduced later.

In the context of modern payment systems, the occurrences of payment primitives (events) change the behaviors of
monetary units in how much monetary values it has over continuous. The building blocks of the financial contracts should
be about modeling these events and their reactivity declaratively.

@@

We also want a formal and precise specification of payment primitives. The title of the paper contains
\textit{denotational semantics}: ``it is a \textit{compositional} style for precisely specifying the meanings of
languages, invented by Christopher Strachey and Dana Scott in the 1960s (\cite{scott1971toward})'', and Conal Elliott
proposed that denotational semantics can also be applied to \textit{data types} in programming \footfullcite[][section
    2, denotational semantics and data types]{Elliott2009-type-class-morphisms-TR}. We will revisit this formulation
too.

@@

@@
